Type-Safe Modular Parsing

### 2017-01-13 paper/sections/Introduction.tex
More Intro

The quest for improved modularity, variability and extensibility of
programs has been going on since the early days of Software
Engineering~\cite{McIlroy68}. Modern Programming Languages enable a certain
degree of modularity, but they have limitations as illustrated by
well-known problems such as the Expression Problem~\cite{wadler1998expression}. The
Expression Problem refers to the difficulty of writing data
abstractions that can be easily extended with both new operations and
new data variants. Traditionally the kinds of data abstraction found
in functional languages can be extended with new operations, but
adding new data variants is difficult. The traditional object-oriented
approach to data abstraction facilitates adding new data variants
(classes), while adding new operations is more difficult.

@@

Semantic modularization techniques go one step further in terms of modularity,
and also enable components or features to be modularly type-checked
and separately compiled. Modular type-checking and separate
compilation are desirable properties to have from a software
engineering point-of-view. Modular type-checking can report errors 
earlier and in terms of the modular code programmers have written 
in the first place. Separate compilation avoids global compilation
steps, which can be very costly. Furthermore semantic modularization 
enables the composition of compiled binaries as well as ensuring the 
type-safety of the code composed of multiple components. Examples of semantic modularization techniques 
include various approaches to \emph{family polymorphism}~\cite{ernst01FP},
virtual classes~\cite{Ernst:2006}, as 
well as several other techniques for solving the Expression
Problem~\cite{Oliveira:2012}\bruno{more here; yanlin; Torgersen; Odersky...}.

Semantic modularization techniques are less widely used in practice 
than syntactic techniques. This is partly due to the perceived need
for sophisticated type systems, which are not available in mainstream 
languages and may require more knowledge from users. However, recently 
several lightweight modularization techniques have been shown to work 
in mainstream programming languages like Java or Scala. Object
Algebras~\cite{Oliveira:2012} are one such technique, which 
Most semantic
modularization techniques have focused on operations that traverse or 
or process extensible datastructures, such as ASTs.
\bruno{Object Algebras here?} However, as far as
we know there is little work on operations that build/produce ASTs. 
In particular the problem of modular parsing has not been studied in
semantic modularization approaches. This is a shame because, to
realise the vision of software product-lines for programming
languages, modular parsing is a necessity.

\paragraph{Modular and Extensible Parsing}
  This paper investigates presents \name: a parsing
  combinator library that enables modular parsing.
  \name provides a solution for the problem of \emph{semantic modular
    parsing}. That is, the solutions should not only
  allow complete parsers to be built out of modular parsing
  components, but also enable the parsing components to be \emph{modularly
  type-checked} and \emph{separately compiled}. \name is
  built on top of a Packrat parsing library, but adds new parsing
  combinators to enable modular parsing. The new parsing combinators 
  employ \emph{delegation-based} techniques and \emph{Object Algebras} 
  to support extensibility. The choice of Packrat parsing over other
  parsing techniques turns out to be important for achieving
  performance in a modular setting. \bruno{left recursion, and
    backtracking removal.}

@@

\item {\bf Algorithmic challenges:} answer Packrat parsing

\item {\bf Typing and Reusability Challenges:} answer Multiple
  inheritance and Object Algebras.

@@

>>>>

The quest for improved modularity, variability and extensibility of
programs has been going on since the early days of Software
Engineering~\cite{McIlroy68}. Modern Programming Languages (PLs) enable a certain
degree of modularity, but they have limitations as illustrated by
well-known problems such as the Expression Problem~\cite{wadler1998expression}. The
Expression Problem refers to the difficulty of writing data
abstractions that can be easily extended with both new operations and
new data variants. Traditionally the kinds of data abstraction found
in functional languages can be extended with new operations, but
adding new data variants is difficult. The traditional object-oriented
approach to data abstraction facilitates adding new data variants
(classes), while adding new operations is more difficult.

@@

Semantic modularization techniques go one step further in terms of modularity,
and also enable components or features to be modularly type-checked
and separately compiled. Modular type-checking and separate
compilation are desirable properties to have from a software
engineering point-of-view. Modular type-checking can report errors 
earlier and in terms of the modular code programmers have written 
in the first place. Separate compilation avoids global compilation
steps, which can be very costly. Furthermore semantic modularization 
enables the composition of compiled binaries as well as ensuring the 
type-safety of the code composed of multiple components. Examples of semantic modularization techniques 
include various approaches to \emph{family polymorphism}~\cite{ernst01FP},
\emph{virtual classes}~\cite{Ernst:2006}, as 
well as several other techniques for solving the Expression
Problem~\cite{Oliveira:2012}\bruno{more here; yanlin; Torgersen; Odersky...}.

Semantic modularization techniques are less widely used in practice 
than syntactic techniques. This is partly due to the need
for more sophisticated type systems, which are not available in mainstream 
languages and may require more knowledge from users. However, recently 
several lightweight modularization techniques have been shown to work 
in mainstream programming languages like Java or Scala. Object
Algebras~\cite{Oliveira:2012} are one such technique, which works in
Java-like languages and uses simple generics only. So far research on semantic
modularization techniques have focused on operations that \emph{traverse} or 
or \emph{process} extensible datastructures, such as ASTs. Indeed 
many documented applications of semantic modularization techniques
focus on modularizing various aspects of PL implementations.
However, as far as we know there is little work on operations that build/produce ASTs. 
In particular the problem of how to modularize parsing has not been studied in
semantic modularization approaches. This is a shame because parsing
is a fundamental part of PL implementations, and it ought to be
modularized as well.

%\paragraph{Modular and Extensible Parsing}

This paper investigates presents a technique for doing semantically
modular parsing.  That is, our approach not only allows complete
parsers to be built out of modular parsing components, but also enables
those parsing components to be \emph{modularly type-checked} and
\emph{separately compiled}. In developing our techniques we encontered 
two different classes of challenges:

@@

\item {\bf Algorithmic challenges:} A first challenge was todo with
  the parsing algorithms themselves. Since many parsing formalisms
  were not designed with extensibility in mind, some of the employed 
  techniques were non-modular, or performed poorly in a modular setting.

\item {\bf Typing and Reusability Challenges:} The second class of
  challenges were problems related to modularity, reusability and
  typing.

@@

\name is built on top of a Packrat parsing
library, but adds new parsing combinators to enable modular
parsing. The new parsing combinators employ \emph{delegation-based}
techniques and \emph{Object Algebras} to support extensibility. The
choice of Packrat parsing over other parsing techniques turns out to
be important for achieving performance in a modular
setting. \bruno{left recursion, and backtracking removal.}
